---
title: "Weather Analysis to Identify Severe Weather Events Affects Population Health and Economic Conditions across U.S"
author: "Abhinav Khandelwal"
---
## Synopsis

Storms and other weather conditions severe affect the population health as well as the economic conditions. This can results in severe Fatalities, Injuries and Property Damages which becomes very important to control these events. 

An analysis has been done on U.S. National Oceanic and Atmospheric Administration's (NOAA). The database is from 1950 to Nov 2011. In this analysis we will identify which weather events are severe to population health and economic condition across U.S.

We have used barplot to plot the data and identified that **Tornado** affects the population health majorly and **Flood** causes major Economic Damage across U.S.

## Data Loading and Processing
```{r}
## Loading required packages
library(dplyr)

## Changing the size of digits
options(scipen = 999,digits = 10)

```

```{r,cache=TRUE}
## Loading the data in R

weather_data = read.csv("C:\\Users\\Abhinav Khandelwal\\Desktop\\R_project\\R_Learning_Coursera\\StormData.csv")

head(weather_data,2)
```

For the analysis i need only few columns like "EVTYPE","FATALITIES","INJURIES","PROPDMG","
PROPDMGEXP","CROPDMG","CROPDMGEXP". I will use Select function to extract these columns.

```{r}
weather_data_cleaned = select(weather_data,"EVTYPE","FATALITIES","INJURIES",contains("DMG"))
## Making all the EVTYPE's in upper case

weather_data_cleaned$EVTYPE = toupper(weather_data_cleaned$EVTYPE)
head(weather_data_cleaned,2)

```

In the dataset the EVTYPE is in character datatype need to change into factor for the analysis

```{r}
weather_data_cleaned$EVTYPE = as.factor(weather_data_cleaned$EVTYPE)
str(weather_data_cleaned)

```

As we can see from above that we have 985 EVTYPE. From the str summary you can see that there are 0's at many places which is not required in our analysis. So we will keep only those rows which will have values.And for the analysis i need to divide the dataset into 2 parts first is related to Population Health and second related to Economic Consequences.

```{r}
popdata = weather_data_cleaned %>% select(1,2,3) %>% filter(FATALITIES > 0 | INJURIES > 0)
head(popdata)

econodata = weather_data_cleaned %>% select(1,contains("DMG")) %>% filter(PROPDMG >0 | CROPDMG > 0)
head(econodata)

```

Let's see the econodata what all EXP we have in PROP and CROP.
```{r}
unique(econodata$PROPDMGEXP)
unique(econodata$CROPDMGEXP)
```
From above, it is quite visible that there are some EXP which are not valid. Valid values are K for kilos, M for millions ,B for billions. 

Replacing all EXP which are invalid to 0 and keeping the valid one's.
```{r}

econodata = econodata %>% mutate(CROPDMGEXP =  case_when(CROPDMGEXP %in% c("K","m","M","B")~CROPDMGEXP,TRUE ~ "0"))

econodata = econodata %>% mutate(PROPDMGEXP =  case_when(PROPDMGEXP %in% c("K","k","b","m","M","B")~PROPDMGEXP,TRUE ~ "0"))

```

Now we will substitute the actual values of EXP in the Econodata.

```{r}
econodata = econodata %>% mutate(PROPEXP = case_when(PROPDMGEXP == "K" ~ PROPDMG * 1000,PROPDMGEXP == "k" ~ PROPDMG * 1000,PROPDMGEXP == "m" ~ PROPDMG * 1000000,PROPDMGEXP == "M" ~ PROPDMG * 1000000,PROPDMGEXP == "B" ~ PROPDMG * 1000000000,PROPDMGEXP == "b" ~ PROPDMG * 1000000000, TRUE ~ PROPDMG))

head(arrange(econodata,desc(PROPEXP),10))

econodata = econodata %>% mutate(CROPEXP = case_when(CROPDMGEXP == "K" ~ CROPDMG * 1000,CROPDMGEXP == "k" ~ CROPDMG * 1000,CROPDMGEXP == "m" ~ CROPDMG * 1000000,CROPDMGEXP == "M" ~ CROPDMG * 1000000,CROPDMGEXP == "B" ~ CROPDMG * 1000000000,CROPDMGEXP == "b" ~ CROPDMG * 1000000000, TRUE ~ CROPDMG))

head(arrange(econodata,desc(CROPEXP),10))
```

Now the cleaning part is done. We will now plot the datasets to answer the questions.

## Results

#### **Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**

Aggregating the Fatalities and Injuries across the EVTYPES and we will keep only those rows where the value of FATALITIES and INJURIES arae more than 1000 otherwise our plot will be messy.

```{r}
popdata_total = popdata%>% group_by(EVTYPE) %>% summarise(sum(FATALITIES),sum(INJURIES)) %>% rename(FATALITIES = "sum(FATALITIES)",INJURIES = "sum(INJURIES)") %>% filter(FATALITIES > 1000 | INJURIES > 1000)%>% arrange(desc(FATALITIES),desc(INJURIES)) %>%as.data.frame()

head(popdata_total)
```

**Plotting the graph which shows which EVTYPES are more harmful to population health**

We will now plot popdata_total to see which weather event is more harmful to population health.

```{r}
barplot(as.matrix(t(popdata_total[,-1])),beside = TRUE,names.arg = popdata_total$EVTYPE,legend.text = c("FATALITIES","INJURIES"),col = c("grey","lightblue"),ylab = "No. of Injuries / Fatalities",las = 3,cex.axis = 01,cex.names = 0.6,main = "Analysis of Events \nwhich are Harmful Respect to Population Health")

```

From the above plot we can conclude that **Tornado** affects the population health majorly.

#### **Across the United States, which types of events have the greatest economic consequences?**

```{r}
econodata_total = econodata %>% group_by(EVTYPE) %>% summarise(sum(PROPEXP),sum(CROPEXP)) %>% rename(Property_Damage = "sum(PROPEXP)",Crop_Damage = "sum(CROPEXP)") %>% arrange(desc(Property_Damage),desc(Crop_Damage)) %>% as.data.frame()
head(econodata_total,10)

## As we have more than 390 observations which will be very difficult to see in the plot so we will keep only those records which will have damages in billions of dollars.

econodata_total = econodata_total %>% filter(Property_Damage > 10000000000 | Crop_Damage > 1000000000)

head(econodata_total)
```

**Plotting the graph which shows which EVTYPES are more harmful to economic consequences**

We will now plot econodata_total to see which weather event caused more harm to economic condition in USA. As the values of the damage are bigger so i have divided the values with 1000000000 so the y-axis is in billion dollars.

```{r}
barplot(t(econodata_total[,-1]/1000000000),names.arg = econodata_total$EVTYPE,las = 3,cex.names = 0.6,cex.axis = 0.7,col = c("brown","grey"),legend.text = TRUE,ylab = "Total Damages in Billion Dollars",main = "Analysis of Events which causes Economic Damage")

```

From the above we can conclude that Flood caused more Economic Damages (Property and Crop Damages)

## Conclusion

From the above plot's we can conclude that **Tornado** weather event majorly affects Population Health in U.S. and **Flood** weather event caused major damage to Economic Conditions across U.S.
